---
title: "Stats Distribution"
author: "Pranav Vogeti"
date: "2023-03-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate) # to extract the year
```
```{css, echo=FALSE}
body {
  background-color: papayawhip !important
}
```

## Objective

The objective of this is to analyze the distribution of Kansas's Basketball stats over the course of the past decade against the rest of the nation by conference categorizations. 


# The Data

Here we read the data for the past several seasons of college basketball. It should be noted that cbb13 indicates the 2012-2013 season, cbb14 indicates the 2013-2014 season, etc. 
```{r read-data}
cbb13 <- read.csv("../data/cbb13.csv")
cbb14 <- read.csv("../data/cbb14.csv")
cbb15 <- read.csv("../data/cbb15.csv")
cbb16 <- read.csv("../data/cbb16.csv")
cbb17 <- read.csv("../data/cbb17.csv")
cbb18 <- read.csv("../data/cbb18.csv")
cbb19 <- read.csv("../data/cbb19.csv")
cbb20 <- read.csv("../data/cbb20.csv")
cbb21 <- read.csv("../data/cbb21.csv")
```
## First Filter

Here we will attempt to obtain the average seed obtained by conference for each season. For example, for the cbb13 season, what was the Big 12's average seed for the teams that earned a post season bid? What about for the SEC? The output should be a table containing three columns:

* `conference` the college basketball conference
* `avgSeed` the average seed by each conference
* `bids` the number of bids that team sent

```{r filter-data-cbb13}
conference_13 <- cbb13 %>%
  mutate(year = as.character("2012-2013")) %>%
  select(CONF, SEED, year) %>%
  rename(conference = CONF, seed = SEED) %>%
  drop_na()
 

##print out the first 10 rows of the 68 row data set
head(conference_13, 10)


##now we add our new columns to our summary table
seeds_13 <- conference_13 %>%
  group_by(conference, year) %>%
  summarize(avgSeed = mean(seed), bids = n()) %>%
  arrange(avgSeed)
 

seeds_13
```

## Using Functions

We can simply copy and paste this set of instructions for each data file, but we can also attempt to do it by defining a function that can do that, thereby creating cleaner code in the process. 

We make a function `seed_breakdown()` that will do exactly what we did with our `cbb13` data set. This function also includes a `mutate()` and a new variable `year` to help characterize our data.

* `year` a new column variable indicating the basketball season

```{r seed-breakdown-function}
seed_breakdown <- function(dataframe, yearString) {
  ##Input: dataframe = already read dataframe
  ##Output: nicely formatted tibble
  
seed_df <- dataframe %>%
  select(CONF, SEED) %>%
  rename(conference = CONF, seed = SEED) %>%
  drop_na() %>%
  group_by(conference) %>%
  summarize(avgSeed = mean(seed), bids = n()) %>%
  arrange(avgSeed) %>%
  mutate(year = as.character(yearString))

return (seed_df)
}
```

## Using our New Function

We can now output tibbles for our stored csv files at our will in a clearer, concise manner thanks to our `seed_breakdown()` function. We will output them separately to see the effect of each function call.

* **NOTE:** We chose to omit `cbb20` simply because there was no postseason due as a consequence of COVID-19

```{r}
##2012-2013 season
cbb13 <- seed_breakdown(cbb13, yearString = "2012-2013")
cbb13

##2013-2014 season
cbb14 <- seed_breakdown(cbb14, yearString = "2013-2014")
cbb14

##2014-2015 season
cbb15 <- seed_breakdown(cbb15, yearString = "2014-2015")
cbb15

##2015-2016 season
cbb16 <- seed_breakdown(cbb16, yearString = "2015-2016")
cbb16

##2016-2017 season
cbb17 <- seed_breakdown(cbb17, yearString = "2016-2017")
cbb17

##2017-2018 season
cbb18 <- seed_breakdown(cbb18, yearString = "2017-2018")
cbb18

##2018-2019 season
cbb19 <- seed_breakdown(cbb19, yearString = "2018-2019")
cbb19

##2020-2021 season
cbb21 <- seed_breakdown(cbb21, yearString = "2020-2021")
cbb21
```

## Creating the Full Data Set

Before we begin doing so, we must join all of our data together so we have one data set we can get our summary from. We will adopt a `full_join()` so that we keep all the data that will be joined and therefore keep all conferences as well, regardless of if they made the tournament or not. 

```{r eval = TRUE}
full_seed_data <- cbb13 %>%
  full_join(cbb14, by = "conference") %>%
  full_join(cbb15, by = "conference") %>%
  full_join(cbb16, by = "conference") %>%
  full_join(cbb17, by = "conference") %>%
  full_join(cbb18, by = "conference") %>%
  full_join(cbb19, by = "conference") %>%
  full_join(cbb21, by = "conference")


full_seed_data  
```

Now we just rename our columns to make it more readable.

```{r eval = TRUE}

full_seed_data <- full_seed_data %>%
  rename(
    avgSeed = contains("avgSeed"),
    bids = contains("bids"),
    year = contains("year")
  ) 

##renamed set
full_seed_data
```



## Extraction for Graphing

I ran into trouble into graphing the data, so I am attempting this extraction as a workaround since we have many duplicate columns and my goal is to graph all of these one plot per category. 

* `years` a vector containing the years (seasons)
* `bids` a vector containing the bids by conference
* `avgSeeds` a vector containing the avgSeeds by conference

```{r eval = TRUE}

#extract the bids as a separate vector from the data table
bids <- full_seed_data %>% 
  select(contains("bids"))
bids

#extract the avgSeeds as a separate vector from the data table
avgSeeds <- full_seed_data %>% 
  select(contains("avgSeed"))
avgSeeds

#extract the years as a separate vector from the data table
years <- full_seed_data %>% 
  select(contains("year"))
years <- years[1,]
years

#extract the conferences as a vector
conferences <- full_seed_data %>% 
  select(conference)
conferences
```

# The Graphs!

Now we can begin graphing! We can graph them all at once or look at these conferences by sub-conferences, if you will (Power 5, Mid-Major, etc.)

## Sub Dividing 

```{r sub-divide, eval = TRUE}
power5 <- full_seed_data %>%
  filter(
    conference == "B10" |
    conference == "B12" |
    conference == "ACC" |
    conference == "SEC" |
    conference == "P12" 
  )

power5

mid_major <- full_seed_data %>%
  filter(
    conference != "B10" &
    conference != "B12" &
    conference != "ACC" &
    conference != "SEC" &
    conference != "P12" 
  ) 

mid_major
```

## avgSeeds vs Years
```{r}
ggplot(avgSeeds, aes(x = conference, y = c(contains("avgSeed")))) + 
  geom_line(aes(col = conferences)) + 
  geom_point() +
  xlab("Year") +
  ylab("Average Seed") +
  ggtitle("Average Seed by Conference", 
          subtitle = "2013-2019 and 2020-21")
```



## bids vs years
```{r}
ggplot(power5, aes(x = years, y = bids)) + 
  geom_line(aes(col = conferences)) + 
  geom_point() +
  xlab("Year") +
  ylab("Number of Bids") +
  ggtitle("Number of Bids Sent by Conference", 
          subtitle = "2013-2019 and 2020-21")
```

