---
title: "Kansas Basketball Points"
author: "Pranav Vogeti"
date: "2023-02-03"
output: html_document
---

```{r message = FALSE, include = FALSE}
# notice this chunk does not show up in knitted html file
library(tidyverse)
library(lubridate) # to extract the year
library(rvest) # this is for the read_html and html_table functions. May have to install it.
```

Objective: To create a visual for looking at Kansas's point totals as a  trend over the course of the regular season to date. 



We begin by feeding in the link from sports reference which contains the data set we will use to get points scored per game over the course of the season.
```{r get-raw-data}
## obtain the raw data from which we will obtain the points
ku_raw_data = html_table(read_html("https://www.sports-reference.com/cbb/schools/kansas/2023-gamelogs.html"))

##output the raw data
ku_raw_data
```

We now have the raw data. But it is currently in the form of a list. The first entry in this list provides us a 34 x 40 tibble containing all of the information we have about Kansas's games for the 2022-2023 season. So we must retrieve this first entry. 

```{r fetch-game-data}
##index to the first entry of this list to retrieve the current season game log
ku_game_log <- ku_raw_data[[1]]

##output the game log
ku_game_log
```

Now we just have the 34 x 40 tibble. From here we can filter out what we want. But there is just one noticeable issue: our first row contains our column headers. The end result should just be a table of the game number and the points scored for that game, and the table should contain appropriate column headers. So, we set row 1 (the row that contains the column headers) as the column headers for the game log, and then we filter our data to our desired variables accordingly. 

```{r filter-game-data}
#sets the first row of the tibble (a 1 x 40 tibble vector) to be the column headers for our data. 
colnames(ku_game_log) = ku_game_log[1,]


##print
ku_game_log

 
##create the game-point log
ku_point_game_log <- ku_game_log %>% 
  ##look for only the columns with the 'G' and 'Tm' column heading
  select(G, Tm, 'W/L') %>%
  ##rename the 'G' column heading as the "Game" column heading
  rename("Game" = G, "Points" = Tm) %>%
  ##every 20 games, sports reference adds a separator row with empty data
  ##We filter out any instance where we have a row containing empty data
  ##We also filter out any row instance where we have the "Tm" column heading"
  filter(Points != "" & Points != "Tm") %>%
  ##currently, the points are of type character. We want them as numbers to fiddle with. 
  mutate (
   Points = as.numeric((Points)),
   Game = as.numeric((Game))
  )
##output the newly filtered data
head(ku_point_game_log)
```

Now we have our point-game log, so we begin graphing!

```{r graphs}
#scatterplot of the points scored
ggplot(ku_point_game_log, 
       aes(x = Game, y = Points)
       ) + geom_point() + geom_vline(xintercept = 13, color = "red", linetype = "dotted")


#line graph of the points scored
ggplot(ku_point_game_log, aes(x = Game, y = Points)) + geom_line() + geom_point() +
geom_vline(xintercept = 13, color = "red", linetype = "dotted")

#NOTE: the red dotted line indicates when conference play has begun

```

A more informative graph would also contain information about how these points occurred. For this case, we decide to also color coordinate the wins and losses for the data set. 

-Let's first create the data set to deal with

```{r win-loss-set creation}
colnames(ku_game_log) = ku_game_log[1,]

ku_win_loss_points <- ku_game_log %>%
  select(G, Tm,"W/L") %>%
  rename("Game" = G,
         "Points" = Tm, 
         "Result" = "W/L") %>%
  filter(Points != "" & Points != "Tm") %>%
  mutate(
    Game = as.numeric(Game),
    Points = as.numeric(Points),
   )
head(ku_win_loss_points)
```

Now we have our new set of data, we can now graph this set!

```{r new-graphs}
ggplot(ku_win_loss_points, aes(x = Game, y = Points)) + 
  geom_line(aes(color = "white")) + 
  geom_point(aes(color = str_sub(Result, 1, 1)), size = 4) +
  geom_vline(xintercept = 13, color = "pink", linetype = "dotted") +
  scale_color_manual(values = c("W" = "yellow", "L" = "red"), name = "test title") +
 theme(panel.background = element_rect(fill = '#082557', color = 'white')) +
  xlab("Game Number") +
  ylab("Points Scored") +
  ggtitle("Kansas Basketball: Points Per Game", 
          subtitle = "2022-2023 Season" )
  
```




