---
title: "Kansas Win-Loss Performances"
author: "Pranav Vogeti"
date: "2023-03-15"
output: html_document
---

```{css, echo=FALSE}
body {
  background-color: papayawhip 
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate) # to extract the year
library(kableExtra) # this is to format our outputted tables
```

# Objective: 

After the rather abysmal showing by the Jayhawks against Texas in the 2023 Big 12 Championship, and after the scores for each loss, I was curious to see if there are particular stats that can be used to obtain an idea of what went wrong for that loss. This will also serve as a general statistical analysis of how the Jayhawks perform when they win and when they lose. 





# Formatting our Data

I'll begin by first loading in the raw data obtained from Sports Reference^[https://www.sports-reference.com/cbb/schools/kansas/men/2023-gamelogs.html]
```{r get-ku-data}
## obtain the raw data from which we will obtain the points
ku_raw_data <- read_csv("../data/kumbb23.csv")

##print out this data
ku_raw_data
```



Before we can truly begin our data adventure, this tibble actually "hides", if you will, the column headers we should be using. Currently, we have character-valued column headers that R made by default. But row 1 of this data contains the column headers we want.

* I'll use `colnames()` to store this row and set it to be the column names for `ku_raw_data` and `subset()` to remove the first row from this dataset

**NOTE:** Credit for this column rename idea goes to my TA for my STAT 240 class. He showed me this process. 
```{r columnnames, eval = TRUE}
##This will set the first row (a 1x40 tibble vector) to be our column header
colnames(ku_raw_data) = ku_raw_data[1,]
## Avoids the duplicate 'Opp' column header and we make the distinction
colnames(ku_raw_data)[3] = "Team Name"

## keep all rows where Tm has a numeric value
ku_true_data <- ku_raw_data %>% subset(Tm != "Tm")

## This will remove the opponent columns from this data set
ku_true_data <- ku_true_data %>% select(-c(23:39))


#print out the formatted data
ku_true_data
```




# Filtering our Formatted Data

Now we can finally fiddle around with our data! I will make use of extensive `dplyr` functions to help filter the data as appropriate. The end goal will be to obtain a summary of the basic stats for only KU, including their wins and losses 

* Basic Stats are defined as
  + W/L as `Result`
  + Points as `PTS`
  + Shooting Percentages 
    + Field Goal Percentage as `FGP`
    + Three Point Field Goal Percentage as `ThrP`
    + Free Throw Percentage as `FTP`
  + Rebounds 
    + Offensive Rebounds as `ORB`
    + Defensive Rebounds as `DRB` calculated from `TRB` - `ORB`
  + Assists as `AST`
  + Steals as `STL`
  + Blocks as `BLK`
  
```{r summaries, eval = TRUE}
##rename and add stat columns
ku_summary <- ku_true_data %>% 
  rename(
    Result = 'W/L',
    PTS = Tm,
    FGP = 'FG%', 
    ThrP = '3P%',
    FTP = 'FT%'
  )  %>%
  mutate(
    ## make everything numeric so that we can fiddle with it
    PTS = as.numeric(PTS),
    FGP = as.numeric(FGP),
    ThrP = as.numeric(ThrP),
    FTP = as.numeric(FTP),
    ORB = as.numeric(ORB),
    TRB = as.numeric(TRB),
    AST = as.numeric(AST),
    STL = as.numeric(STL),
    BLK = as.numeric(BLK)
  ) %>%
  mutate(
    #add the defensive rebound column
    DRB = TRB - ORB
  ) %>%
  select(
    ## keep the stat columns that we want
    Result, PTS, AST, TRB, DRB, ORB, STL, BLK, FGP, ThrP, FTP
  )


##print out the first 10 rows of this summary data
print(ku_summary, n = 10)
```





## Summarization

In this section, I am attempting to summarize these stats by wins and losses. In other words, what are the averages of each of these stat markers when Kansas wins? Loses?

I will use `group_by()` alongside `summarize()` to obtain this information

* The averages are defined by the convention: `avgSTAT`
* The summarized table will be formatted using `kable_styling()` 
```{r super-summary}

ku_super_summary <- ku_summary %>%
  group_by(Result) %>%
  summarize(
    avgPTS = mean(PTS),
    avgAST = mean(AST),
    avgSTL = mean(STL),
    avgBLK = mean(BLK),
    avgTRB = mean(TRB),
    avgDRB = mean(DRB),
    avgORB = mean(ORB),
    avgFGP = mean(FGP),
    avgThrP = mean(ThrP),
    avgFTP = mean(FTP)
  )

#ku_super_summary <- ku_super_summary %>%
 #pivot_longer(
    #cols = c("avgPTS", "avgAST", "avgSTL", "avgBLK", "avgTRB", "avgDRB", "avgORB", "avgFGP", "avgThrP", "avgFTP"),
    #names_to = c(".value", "Averages"), 
    #names_sep = "avg")

#put the data in a kable-styled table
ku_super_summary %>% 
  kable() %>% 
  kable_styling(
    position = "left", 
    full_width = FALSE,
    bootstrap_options = c("striped")
  )
```

## Data Analysis

Based on this data, it seems that the main difference in a Win Loss results in ball movement. Excluding OT cases, Kansas averages 11.2 assists in a loss against 17.5 assists in a win. Those six assists definitely impact point production as evidenced by the point differences. Moreover, every loss, they've faced staunch defenses as their Field Goal Percentage dips from 48.6% to 38.9% by loss, a near 10% decrease. This magnitude of a decreases is also seen in three point percentage, from 36.5% to 27.5%. The assist column demonstrates that when the team passes the ball more, more open looks are generated, explaining these field goal percentage disparities. 

By this data alone, Kansas loses when they do not move the ball, or when they encounter a defense that forces isolations, stifling stars from open court movement. 

# Conclusions and Extensions

Overall, this was a fun project to work on. I actually tried to import the data directly from a link and use `rvest` library to unravel it but I encountered empty values and duplicate column names that I just could not work around. This project introduced me to how to import csv text to an Excel file and convert it to a csv for simplicity. I also was able to use the `kableExtra` library to format my outputted summary data which I now know I will make extensive use of. 

As an extension of this project, I do want to learn to pivot this data to tidy format to visually graph this data. I hope to this in the very near future.  

Another extension that would be interesting is to look at an opponent that Kansas has lost to frequently (i.e Texas) and investigate if their win/loss numbers correlate to the scores presented. 


# References