---
title: "NBA and the Mid-Range"
author: "Pranav Vogeti"
date: "2024-01-07"
output: 
  html_document:
    css: customOutputScript.css
    toc: true
    toc_float: true
---

```{r message = FALSE, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(lubridate)
library(kableExtra)
library(sportyR)
```


# Introduction

As we inch closer to the NBA All-Star Game, a game of absent defense and HIGH scoring offense, I was interested in analyzing how the Mid-Range shot in the NBA has evolved over time.


### The Focus Question

How has the mid-range shot in the NBA changed over time, and are there any influences that have mediated this particular shot's volume?

### Motivation

There are many different types of offensive plays in the NBA. Some plays are designed for guards to drive the ball into the basket. Some plays are designed for a quick three point shot. Some plays are designed to wear out the defense before making a decision on what shot to take. 

There are different types of shots in an NBA game. The dunk, the layup, paint shots, the free throw, the three point shot, and the focus of this report: the mid-range. In the early years of the modern NBA (80s-90s), the three point shot was more or less a privilege and an advantage for the undersized guard. Offenses at that time were largely based on physicality and paint-presence. Not many three point shots were taken and were treated as fall back options. Interestingly, if one were to turn on an NBA game now, that is not the case. Even the NBA's big men have developed a competitve jump shot to add an extra dynamicism to their offensive capabilities. Players like Nikola Jokić, Karl-Anthony Towns, Brook Lopez, Joel Embiid, and many other counterparts are prime examples of this evolution. In fact, in the 2022 three point contest, it was the big man Karl-Anthony Towns that won the infamous shooting contest! Towns was only the third big man to do so, behind very famous players like Dirk Nowitzki and Kevin Love. [^1]

[^1]: https://sports.yahoo.com/minnesota-timberwolves-center-karl-anthony-towns-wins-nba-all-star-3-point-contest-025319348.html

The three point shot has become so commonplace that it is hard to find plays that use the mid-range. If the league's big men have credible three-point shots, defenses get spread out in the half-court as now every offensive player can get a three over a two. A spread out defense also leaves gaping holes that, when properly accounted for by the offense, allows guards to simply drive the ball to the paint for a layup, or dunk, or a chance at the free-throw line. Thus, it is worth investigating the presence (or lackthereof, potentially) of the mid-range and its relations to the three point shot. 

# Background

### About the Original Data: Play-by-Play

The author that has provided the data for us to use is Vladislav Shufinsky. He has compiled NBA play-by-play data from 1996-2022 and made it open-source on Kaggle to allow us to use it for excursions like this[^2].

[^2]: <https://www.kaggle.com/datasets/brains14482/nba-playbyplay-and-shotdetails-data-19962021/data>

The data set we will use will contain a whole season's log of play-by-play data. In order to see changes over time, we will use the collection available from this data set. This collection spans from the 2015-2016 NBA season to the 2021-2022 NBA season inclusive. While the entire dataset consists of four subdata sets, we will choose to use the data originally sourced from *data.nba.com* This means that we will use the files called "*datanba_YYYY.csv*" where YYYY is the season of choice. So, the 2015-2016 season uses the "*datanba_2016.csv*

### The Key Variables

-   The key variables from the *Play-By-Play* we will use for this excursion:
    -   `cl` the time stamp at which the significant play occurred
    -   `de` the short form description of what the play was
    -   `locX` the x-coordinate location of the `de` play
    -   `locY` the y-coordinate location of the `de` play
    -   `hs` the home team total points at the `cl` moment
    -   `vs` the visiting team total points at the `cl` moment
    -   `PERIOD` the quarter of the game so as to distinct `cl`
    -   `GAME_ID` the numerical identifier of the game being played

### Assumptions and Unusuality

A giant aspect of this report relies on the location data: `locX` and `locY`. The way we define the type of shot performed agrees with the NBA standard. However, there are certainly edge cases where we classify a shot differently when compared to the actual scorekeeper's book. Such edge cases are rare, but can and will happen during this excursion. An example of this case can be found in Game 6 of the 2021 NBA Eastern Conference Finals, when the Bucks played the Nets. In that game, Kevin Durant took a shot that had merits to be a three, but was ruled as a two because of his shoe being on the three point line. We will interpret all shots that show on the marking lines as a two, even if some of them were ruled as a three. 

Another aspect of this report is that we are not necessarily evaluating the player's position. This is more of a holistic look at the shot architecture rather than a case-by-case shot outlook. So, a shot taken at the three point line by a Center will be treated the exact same as that shot taken by a Point Guard. We are just interested in looking at a broad look at shot behavior without delving too much into particulars. 

### Report Intentions

The rest of this report will focus on cleaning the data to fit our needs, and analyzing said data and fostering meaningful observations and conclusions. 


# Analysis

## Data Cleanup

### Getting the Shots

First we load in the data. 
```{r load-data , echo=T}
# read the play by play data. this is an expensive chunk to load so run sparingly
raw_pbp_2016_data <- read.csv("../data/datanba_2016.csv")
raw_pbp_2017_data <- read.csv("../data/datanba_2017.csv")
raw_pbp_2018_data <- read.csv("../data/datanba_2018.csv")
raw_pbp_2019_data <- read.csv("../data/datanba_2019.csv")
raw_pbp_2020_data <- read.csv("../data/datanba_2020.csv")
raw_pbp_2021_data <- read.csv("../data/datanba_2021.csv")
raw_pbp_2022_data <- read.csv("../data/datanba_2022.csv")

##print the first 15 rows of a sample of this raw data
head(raw_pbp_2022_data, n=15)
```

Now, we will create our cleaned data sets. Since we are dealing with the same type of data, it makes sense to create a function called `get_shots_data()` that will do this for us. 

```{r get-shot-data-function, echo=T}
##the last name of the player, since that is what is in the pbp data
player_regex = "\\]\\s\\w+"
##to capture all "shot" from the play description
shot_literal = "shot"
##to capture all "3pt" shots from the play description
three_shot_literal = "3pt"
##to capture the team code
team_regex = "\\[[A-Z]{3}"
##to capture all "made" shots from the play description
made_literal = "Made"
##to capture all "missed" shots from the play description
missed_literal = "Missed"

get_shots_data <- function(raw_pbp_data, reg_season) {

  ##get the variables we use
  pbp_data <- raw_pbp_data %>% select(
    GAME_ID, PERIOD, cl, de, locX, locY, hs, vs
  )
  
  ##get the shot data
  shots_data <- pbp_data %>% 
    filter(str_detect(de, pattern=shot_literal) | str_detect(de, pattern=three_shot_literal)) %>%
    group_by(GAME_ID) %>%
    ## responsible for getting the teams involved in this game
    mutate(
      team = str_extract(de, regex(team_regex, FALSE)),
      player_team = substring(team, 2, 5), 
    ) %>%
    ##responsible for getting the player shooting the shot
    mutate(
      player_identifier = str_extract(de, regex(player_regex, TRUE)), 
      player_shooting = substring(player_identifier, 3, last = 1000000L)
    ) %>% 
    ##responsible for classifying certain stage of the game in period.int format
    mutate(
      interval = case_when(
        cl <= "01:00" ~ 0.5,
        cl <= "03:00" ~ 0.4, 
        cl <= "06:00" ~ 0.3, 
        cl <= "09:00" ~ 0.2,
        cl <= "12:00" ~ 0.1), 
      stage_of_game = PERIOD + interval 
    ) %>% 
    mutate(
      shot_result = case_when(
        str_detect(string=de, pattern=made_literal) ~ "Made",
        str_detect(string=de, pattern=missed_literal) ~ "Missed"
      )
    ) %>%
    ##QoL change
    rename(gm_quarter = PERIOD) %>%
    ##add the season classifier
    mutate(season = reg_season) %>% ungroup() %>%
    select(season, gm_quarter, stage_of_game, de, player_shooting, player_team, shot_result, locX, locY)
  
  return (shots_data)
}
```

Now we use `get_shots_data()` to get our cleaner shot data for each data set. Note, we are not done cleaning after this, but it is a nice intermediate step to have so that we can avoid running our ` r load_data` chunk. 

```{r get-shots-data, echo=T}
##get the shots taken for the 2021-2022 NBA season
inter_shots_2022 <- get_shots_data(raw_pbp_data = raw_pbp_2022_data, "2021-2022")

##get the shots taken for the 2020-2021 NBA season
inter_shots_2021 <- get_shots_data(raw_pbp_data = raw_pbp_2021_data, "2020-2021")

##get the shots taken for the 2019-2020 NBA season
inter_shots_2020 <- get_shots_data(raw_pbp_data = raw_pbp_2020_data, "2019-2020")

##get the shots taken for the 2018-2019 NBA season
inter_shots_2019 <- get_shots_data(raw_pbp_data = raw_pbp_2019_data, "2018-2019")

##get the shots taken for the 2017-2018 NBA season
inter_shots_2018 <- get_shots_data(raw_pbp_data = raw_pbp_2018_data, "2017-2018")

##get the shots taken for the 2016-2017 NBA season
inter_shots_2017 <- get_shots_data(raw_pbp_data = raw_pbp_2017_data, "2016-2017")

##get the shots taken for the 2015-2016 NBA season
inter_shots_2016 <- get_shots_data(raw_pbp_data = raw_pbp_2016_data, "2015-2016")

##print out a sample of what these data frames look like
head(inter_shots_2017, n=10)
```

```{r combine-shot-data, echo=T}
all_shots_data <- rbind(inter_shots_2016, inter_shots_2017, inter_shots_2018, inter_shots_2019, inter_shots_2020, inter_shots_2021, inter_shots_2022)
```


### Understanding Location Data: $N$

Now that we have all of our shots, it's time to classify each shot. Before we do, we need to understand what `locX` and `locY` are all about. 

`locX` and `locY` are the standard media coordinates used by the NBA teams to locate where a player is with reference to the basketball court. The caveat to this system is that these coordinates represent an NBA-regulation half-court plane. 

For reference, let $x_n$ represent `locX` and $y_n$ represent `locY` such that a player's position $p_n$ on the half-court at any given moment is $p_n = (x_n, y_n)$. The origin of this system is directly underneath the basket. We define this origin as $o_n$ with $o_n = (0, 0)$. 

Let's visualize this through some plotting. We do this in a rather fun way: we will create a function called `graph_player_shots()` that will take in a `player` and a `reg_season` and plot all shots with respect to those parameters. 

```{r plotting-on-N, echo=T}
graph_player_shots <- function(player_name, reg_season) {
  
  ##get the shot data of the player and the season
  player_shots_data <- all_shots_data %>% 
    filter((player_shooting == player_name) & (season == reg_season))
  
  ##generate a plot of this shot data
  player_shot_plot <- ggplot(data = player_shots_data, aes(x=locX, y=locY, color=shot_result)) +
    geom_point(size = 2, alpha = 0.75)  + 
    ylim(-100, 900) + 
    xlab("X-Coordinate Location") + 
    ylab("Y-Coordinate Location") +
    ggtitle(paste(player_name, "'s Shot Location Data"), 
            subtitle = paste(reg_season, "NBA Regular Season")) +
    theme_dark()
  
  ##return this plot
  return (player_shot_plot)
}


##graph Lowry's shots in the 2016-2017 NBA regular season
graph_player_shots(player_name = "Lowry", reg_season = "2016-2017")

##graph Brunson's shots in the 2020-2021 nba regular season
graph_player_shots(player_name = "Brunson", reg_season = "2020-2021")

##graph Lillard's shots in the 2021-2022 nba regular season
graph_player_shots(player_name = "Lillard", reg_season = "2021-2022")

```

Through the above plots, we can see not only every specific shot each player took, but also colored based on whether they made it or not. However, one noticeable problem is standardization. Following our convention, all of these are examples of $p_n$ on $N$, However, the $x_n$ and $y_n$ (recall, $p_n= (x_n, y_n)$) scales are not the same. This something we have to fix, by using a package called `sportyR`. Using this package, we can standardize our $p_n$ to `sportyR`'s coordinate-system making it easy to classify our shots.  

Before we leave this section, let's visualize the numeric bounds of $N$

```{r defining-N, echo=T}
##boundaries of locX and locY which defines N
maxX_N = max(all_shots_data$locX)
minX_N = min(all_shots_data$locX)
maxY_N = max(all_shots_data$locY)
minY_N = min(all_shots_data$locY) 

originX.N = 0
originY.N = 0

##coordinate labels
left_top_N_text = paste("(",minX_N, ",", maxY_N, ")")
right_top_N_text = paste("(", maxX_N, ",", maxY_N, ")")
left_bottom_N_text = paste("(", minX_N, ",", minY_N, ")")
right_bottom_N_text = paste("(", maxX_N, ",", minY_N, ")")
origin_N_text = paste("(", originX.N, ",", originY.N, ")")

## plot our boundaries
ggplot() + 
  geom_point(aes(x=maxX_N,y=maxY_N), colour = "red", size = 4) + 
  geom_point(aes(x=maxX_N,y=minY_N), colour = "red", size = 4) +
  geom_point(aes(x=minX_N,y=maxY_N), colour = "red", size = 4) +
  geom_point(aes(x=minX_N,y=minY_N), colour = "red", size = 4) +
  geom_point(aes(x= originX.N, y=originY.N), colour = "blue", size = 4) +
  theme_dark()
```

The points signify these boundary markers: 

  * Our top left red point has coordinates: `r left_top_N_text[1]`
  * Our top right red point has coordinates: `r right_top_N_text[1]`
  * Our bottom left red point has coordinates: `r left_bottom_N_text[1]`
  * Our bottom right red point has coordinates: `r right_bottom_N_text[1]`
  * The blue point signifies our origin $o_n$ with coordinates: `r origin_N_text[1]`

From this chunk we know the following: 
For `locX` ($x_n$) and `locY` ($y_n$), with respect to a player's position $p_n = (x_n, y_n)$ on the basketball court plane $N$, we define $N$ as follows: 

$$
N = \{p_n = (x_n,y_n) \mid x_n \in [`r minX_N[1]`, `r maxX_N[1]`], y_n \in [`r minY_N[1]`, `r maxY_N[1]`]\}
$$

This can be read as: 

> The court system that the NBA uses for shot-plotting (call it $N$), is simply the whole collection of all possible player positions $p_n$ where $p_n$ is simply a point on $N$ defined by the player's x-coordinate on $N$ (called $x_n$) and the player's y-coordinate on $N$ (called $y_n$) to give us $p_n = (x_n, y_n)$. The interval $x_n \in [`r minX_N[1]`, `r maxX_N[1]`]$ defines all possiblilities that $x_n$ can be on this court $N$. Similarly, the interval $y_n \in [`r maxY_N[1]`, `r maxY_N[1]`]$ defines all possiblilities that $y_n$ can be on this court $N$.


### Understanding Location Data: $R$


As we did for our location data, we defined a coordinate system that we called $N$. Now, let's define a coordinate system for `sportyR`

Additionally, let $x_r$ represent the traditional x coordinate in the basketball court plane that `sportyR` uses for its graphs. Consequently, let $y_r$ represent the traditional y coordinate in the basketball court plane that `R` uses for its graphs. Then, with respect to any traditional basketball court that `sportyR` uses, we define a position $p_r$ as the generalized location of any point on these basketball courts that `sportyR` uses such that $p_r = (x_r, y_r)$. Naturally, this means $p_r$ has origin $o_r = (0, 0)$ which we will establish in the upcoming visualizations. 

Let's visualize $R$

```{r basketball-court-visualization, echo=T}
#court parameters/selectors
LEAGUE = "NBA"
COURT_VIEWS = c( "full", "offense", "offensive_key")
DEG_ROTATE = -90

##generate an NBA regulation offensive-half-court
basketball_court <- geom_basketball(league = LEAGUE, display_range = COURT_VIEWS[1], rotation = DEG_ROTATE)

basketball_court
```

Let's go ahead and identify our boundaries. Since we know all of our shots must be inside the court perimeter, it makes sense to exclude the out-of-bounds region when defining our boundaries. We also know that an NBA court has regulation dimensions: 94 ft X 50 ft. Since, we are dealing with the half-court perimeter, proportionately the half-court has dimensions 47 ft X 50 ft. 

```{r court-boundaries, echo=T}
##boundary constants
LEFT_LINE = -25
RIGHT_LINE = 25
TOP_LINE = 47
BOTTOM_LINE = -47
originX.R = 0
originY.R = 0

##coordinate labels
left_top_coord_text = paste("(", LEFT_LINE, ",", TOP_LINE, ")")
right_top_coord_text = paste("(", RIGHT_LINE, ",", TOP_LINE, ")")
left_bottom_coord_text = paste("(", LEFT_LINE, ",", BOTTOM_LINE, ")")
right_bottom_coord_text = paste("(", RIGHT_LINE, ",", BOTTOM_LINE, ")")
origin_coord_text = paste("(", originX.R, ",", originY.R, ")")

##plot the boundaries
half_court_boundaries <- basketball_court +
  geom_hline(yintercept = TOP_LINE, color = "navyblue", linewidth = 2) +
  geom_hline(yintercept = BOTTOM_LINE, color = "steelblue", linewidth = 2) +
  geom_vline(xintercept = LEFT_LINE, color = "maroon", linewidth = 2) +
  geom_vline(xintercept = RIGHT_LINE, color = "red", linewidth = 2) + 
  geom_point(aes(x=LEFT_LINE, y=TOP_LINE), colour="limegreen", size = 4) + 
  geom_point(aes(x=RIGHT_LINE, y=TOP_LINE), colour="limegreen", size = 4) +
  geom_point(aes(x=LEFT_LINE, y=BOTTOM_LINE), colour="limegreen", size = 4) +
  geom_point(aes(x=RIGHT_LINE, y=BOTTOM_LINE), colour="limegreen", size = 4) +
  geom_point(aes(x=originX.R, y = originY.R), colour = "black", size = 4)
  

## visualize the boundaries
half_court_boundaries
```

Let's quickly discuss what we see. 
The navy blue line is our top border with $y_r = `r TOP_LINE[1]`$.
The steel blue line is our bottom border with $y_r = `r BOTTOM_LINE[1]`$. 
The maroon line is our left border with $x_r = `r LEFT_LINE[1]`$. 
The red line is our right border with $x_r = `r RIGHT_LINE[1]`$. 

The green points signify these line intersections: 

  * Our top left green point has coordinates: `r left_top_coord_text[1]`
  * Our top right green point has coordinates: `r right_top_coord_text[1]`
  * Our bottom left green point has coordinates: `r left_bottom_coord_text[1]`
  * Our bottom right green point has coordinates: `r right_bottom_coord_text[1]`
  
The black point is our origin $o_r$ with coordinates: `r origin_coord_text[1]`


We can now define $R$ more rigorously. For any position $p_r$ that is in $R$, we have that: 

$$
R = \{p_r = (x_r,y_r) \mid 
      x_r \in [`r LEFT_LINE[1]`, `r RIGHT_LINE[1]`], y_r \in [`r BOTTOM_LINE[1]`, `r TOP_LINE[1]`]
\}
$$
This can be read as: 

> The court system that `sportyR` uses for shot-plotting (call it $R$), is simply the whole collection of all possible player positions $p_r$ where $p_r$ is simply a point on $R$ defined by the player's x-coordinate on $R$ (called $x_r$) and the player's y-coordinate on $R$ (called $y_r$) to give us $p_r = (x_r, y_r)$. The interval 
$x_r \in [`r LEFT_LINE[1]`, `r RIGHT_LINE[1]`]$ defines all possiblilities that $x_r$ can be on this court $R$. Similarly, the interval $y_r \in [`r BOTTOM_LINE[1]`, `r TOP_LINE[1]`]$ defines all possiblilities that $y_r$ can be on this court $R$.


### Going from $N$ to $R$ to get $C$. 

Now, we have established two coordinate systems: 

$N$, which is the coordinate system that our data inherently uses, is defined as: 

$$
N = \{p_n = (x_n,y_n) \mid x_n \in [`r minX_N[1]`, `r maxX_N[1]`], y_n \in [`r minY_N[1]`, `r maxY_N[1]`]\}
$$

$R$, which is the coordinate system that `sportyR` uses, is defined as: 

$$
R = \{p_r = (x_r,y_r) \mid 
      x_r \in [`r LEFT_LINE[1]`, `r RIGHT_LINE[1]`], y_r \in [`r BOTTOM_LINE[1]`, `r TOP_LINE[1]`]
\}
$$
To draw attention to our goal, the end goal is to standardize our location data so that we can, under one coordinate system $C$, define what each shot based on the precision of our location data. This will remove any ambiguity on how we define two point shots. Thus, we need to define our new origin $o$ that will be the foundation point that $C$ uses.

Mathematically, we make a general statement that will be more rigorously defined later: 

For standardized $x$ location and standardized $y$ location of a player's position $p$ where $p = (x, y)$), let our final coordinate system $C$ be defined as follows:  

$$
C = \{p = (x,y) \mid x \in \mathbb{R}, y \in \mathbb{R}\}
$$

So, we need to make `locX` ($x_n$) and `loc_y`($y_n$) usable so that we can plot these points $p_n$ on `sportyR`'s court $R$. This will take several translations, hence why we will go through the effort to define one more, final coordinate system that has everything standardized. We call this coordinate system $C$. To get $C$, there are three major steps we need to take: 

    1) Translate $o_n$ to $o_r$, giving us a new origin $o$
    2) Translate $R$ to scale under it's new origin $o$
    3) Define final court coordinate system $C$ under the geometric transitions


#### Translating $o_r$ to $o_n$ to get $o$

To see how far we need to translate $o_n$ to $o_r$, let's first plot $o_n$ on $R$. From [Understanding Location Data: $N$], we know that $o_n$ lies underneath the basket on the TV offensive half-court. 
Let's find what the coordinates of $o_n$ would be then if we plot it underneath the basket of $R$. 

```{r plot-originN-on-R, echo=T}
## tweak these numbers until we find the translation
shift_x = 0
shift_y = -42

## "(x_trans, y_trans)"
shifted_origin_text = paste("(", shift_x, ",", shift_y, ")")


half_court_boundaries +
  geom_point(aes(x=shift_x, y=shift_y), colour="yellow", size = 4) 
```
The above tells us the following about visualizating $o_n$ on $R$: 

> Keeping $y_r \in o_r$ the same ($y_r = 0$), if we move $o_r$ right by 42 $x_r$ units, we can find the origin equivalent of $o_n$ on this system $R$.


#### Using $o$ to redefine $R$

Luckily, after seeing how $o_n$ looked in $R$, we do not have to do too much translation work. 
For one, we were able to keep $y_r \in o_r$ the same ($y_r = 0$).  We just have to shift our $x_r$ rightwards by 42 and keep our $y_r$ the same

Let's define a new basketball court with these shifts in mind
```{r translate-geometries, echo=T}
originX = 0
originY = 0

##new court, rotate -90, so really we need to make an x-translation equivalent to our y-translation from the previous court plot
new_court <- geom_basketball(league=LEAGUE, display_range=COURT_VIEWS[1], rotation=DEG_ROTATE,
                             x_trans = shift_y)

new_court + geom_point(aes(x = originX, y=originY), color = "yellow", size = 4)
```
This has our origin $o$ in yellow with coordinates `origin_coord_text[1]` that agrees with $o_n$ convention. Now, let's redefine $R$. In doing so, we can scale this new $R$ to agree with $N$ and therefore the resulting coordinate system is our final coordinate system $C$. 


```{r new-court-boundaries, echo=T}
##boundary constants
LEFT_X = -25
RIGHT_X = 25
TOP_Y = 42
BOTTOM_Y = -5

##coordinate labels
left_top_text = paste("(", LEFT_X, ",", TOP_Y, ")")
right_top_text = paste("(", RIGHT_X, ",", TOP_Y, ")")
left_bottom_text = paste("(", LEFT_X, ",", BOTTOM_Y, ")")
right_bottom_text = paste("(", RIGHT_X, ",", BOTTOM_Y, ")")
origin_text = paste("(", originX, ",", originY, ")")

new_court_boundaries <- new_court +
  geom_hline(yintercept = TOP_Y, color = "navyblue", linewidth = 2) +
  geom_hline(yintercept = BOTTOM_Y, color = "steelblue", linewidth = 2) +
  geom_vline(xintercept = LEFT_X, color = "maroon", linewidth = 2) +
  geom_vline(xintercept = RIGHT_X, color = "red", linewidth = 2) + 
  geom_point(aes(x=LEFT_X, y=TOP_Y), colour="limegreen", size = 4) + 
  geom_point(aes(x=RIGHT_X, y=TOP_Y), colour="limegreen", size = 4) +
  geom_point(aes(x=LEFT_X, y=BOTTOM_Y), colour="limegreen", size = 4) +
  geom_point(aes(x=RIGHT_X, y=BOTTOM_Y), colour="limegreen", size = 4) +
  geom_point(aes(x=originX.R, y = originY.R), colour = "black", size = 4)

new_court_boundaries
```

Let's quickly discuss what we see. 
The navy blue line is our top border with $y_r = `r TOP_Y[1]`$.
The steel blue line is our bottom border with $y_r = `r BOTTOM_Y[1]`$. 
The maroon line is our left border with $x_r = `r LEFT_X[1]`$. 
The red line is our right border with $x_r = `r RIGHT_X[1]`$. 

The green points signify these line intersections: 

  * Our top left green point has coordinates: `r left_top_text[1]`
  * Our top right green point has coordinates: `r right_top_text[1]`
  * Our bottom left green point has coordinates: `r left_bottom_text[1]`
  * Our bottom right green point has coordinates: `r right_bottom_text[1]`
  
The black point is our origin $o_r$ with coordinates: `r origin_text[1]`

Mathematically, $R$ has been redefined this way: 

$$
R = \{p_r = (x_r,y_r) \mid 
      x_r \in [`r LEFT_X[1]`, `r RIGHT_X[1]`], y_r \in [`r BOTTOM_Y[1]`, `r TOP_Y[1]`]
\}
$$
This can be read as: 

> The court system that `sportyR` uses for shot-plotting (call it $R$), is simply the whole collection of all possible player positions $p_r$ where $p_r$ is simply a point on $R$ defined by the player's x-coordinate on $R$ (called $x_r$) and the player's y-coordinate on $R$ (called $y_r$) to give us $p_r = (x_r, y_r)$. The interval $x_r \in [`r LEFT_X[1]`, `r RIGHT_X[1]`]$ defines all possiblilities that $x_r$ can be on this court $R$. Similarly, the interval $y_r \in [`r BOTTOM_Y[1]`, `r TOP_Y[1]`]$ defines all possiblilities that $y_r$ can be on this court $R$.


#### Defining C

To quickly recap, we have defined a coordinate system called $N$ in [Understanding Location Data: $N$]. We have also defined a separate coordinate system called $R$ in [Understanding Location Data: $R$]. In this section [Going from $N$ to $R$ to get $C$], our aim is to get $C$, a final coordinate system that we can use to classify our shots. 

Since we have $o_n = o_r = `r origin_text[1]`$, all we need to do is apply a set of scalars $k_x, k_y$ to scale $p_r$ to $p_n$. These scaled $p_r * k$ values will then be interpreted as our possible $p$ coordinates for our final coordinate system $C$. 


Now, let's recap our coordinate systems:  

$N$, which is the coordinate system that our data inherently uses, is defined as: 

$$
N = \{p_n = (x_n,y_n) \mid x_n \in [`r minX_N[1]`, `r maxX_N[1]`], y_n \in [`r minY_N[1]`, `r maxY_N[1]`]\}
$$

$R$, which is the coordinate system that `sportyR` uses, has been **NEWLY** redefined as: 

$$
R = \{p_r = (x_r,y_r) \mid 
      x_r \in [`r LEFT_X[1]`, `r RIGHT_X[1]`], y_r \in [`r BOTTOM_Y[1]`, `r TOP_Y[1]`]
\}
$$

Since $N$ has a larger set of bounds than $R$, we define $K$ as the general ratio between these bounds: 
$K = \frac{R}{N}$. In this way, we can use our `locX` ($x_n$) and `locY` ($y_n$) data and apply $K$ on those to make them graph-friendly on $C$. 

For $(x_n, y_n) = p_n \in N$  

Coordinate system $C$ is the result of shrinking $N$'s bounds to agree with $R$'s bounds through 
scalar $K = \{K_x, K_y\}$

$$
C = \{p = (x, y) \mid x = \frac{x_n}{K_x}, y = \frac{y_n}{K_y}\}
$$
For $K_x, K_y \in K$


Let's now implement $K$!

```{r scalar-K, echo=T}
LENGTH.N = abs(maxX_N - minX_N)
WIDTH.N = abs(maxY_N - minY_N)

LENGTH.R = abs(RIGHT_X - LEFT_X)
WIDTH.R = abs(TOP_Y - BOTTOM_Y)

K.x = LENGTH.R/LENGTH.N
K.y = WIDTH.R/WIDTH.N

all_shots_data <- all_shots_data %>%
  mutate(c.x = locX * K.x, 
         c.y = locY * K.y)
```

```{r C-bounds, echo=T}
##boundaries of c.x and c.y which defines C
maxX_C = max(all_shots_data$c.x)
minX_C = min(all_shots_data$c.x)
maxY_C = max(all_shots_data$c.y)
minY_C = min(all_shots_data$c.y)
```

Having added all of our new $p$ to the data set, we have the following mathematical definition for $C$.

$$
C = \{p = (x,y) \mid 
      x \in [`r minX_C[1]`, `r maxX_C[1]`], y_r \in [`r minY_C[1]`, `r maxY_C[1]`]
\}
$$
This can be read as: 

> The final court system that we use for shot-plotting (call it $C$), is simply the whole collection of all possible player positions $p$ where $p$ is simply a point on $C$ defined by the player's x-coordinate on $C$ (called $x$) and the player's y-coordinate on $C$ (called $y$) to give us $p = (x, y)$. The interval 
$x \in [`r minX_C[1]`, `r maxX_C[1]`]$ defines all possiblilities that $x$ can be on this court $C$. Similarly, the interval $y \in [`r minY_C[1]`, `r maxY_C[1]`]$ defines all possiblilities that $y$ can be on this court $C$.

```{r final-court, echo=T}
final_court <- geom_basketball(league=LEAGUE, display_range=COURT_VIEWS[3], rotation=DEG_ROTATE,
                             x_trans = shift_y)
```


### Classifying our Shots

Here is the crux of our data cleanup: classifying our shots. To justify $C$, let's create a function that is very similar to `graph_player_shots()`, with the only change being we use `c.x` and `c.y` instead of `locX` and `locY`






```{r graph-standardized-shots, echo=T}
graph_standardized_shots <- function(player_name, reg_season) {
  
  ##get the shot data of the player and the season
  player_shots_data <- all_shots_data %>% 
    filter((player_shooting == player_name) & (season == reg_season))
  
  ##generate a plot of this shot data using (x,y) E C
  player_shot_plot <- final_court +
    geom_point(data = player_shots_data, aes(x=c.x, y=c.y, color=shot_result)) +
    geom_point(size = 1, alpha = 0.75)  + 
    xlab("X-Coordinate Location") + 
    ylab("Y-Coordinate Location") +
    ggtitle(paste(player_name, "'s Shot Location Data"), 
            subtitle = paste(reg_season, "NBA Regular Season"))
  
  ##return this plot
  return (player_shot_plot)
}


##graph Lowry's shots in the 2016-2017 NBA regular season
graph_standardized_shots(player_name = "Lowry", reg_season = "2016-2017")

##graph Brunson's shots in the 2020-2021 nba regular season
graph_standardized_shots(player_name = "Brunson", reg_season = "2020-2021")

##graph Lillard's shots in the 2021-2022 nba regular season
graph_standardized_shots(player_name = "Lillard", reg_season = "2021-2022")
```



### Analysis 2: 

[Placeholder]

# Discussion

In this section, we will discuss the results of our tests and models and their implications in the context of our question and also their broader implications and limitations. 

### Comparing the Graphs & Overall Conclusions

#### Graph Comparisons Discussion

[Placeholder]

### Shortcomings and Limitations


#### Limitation 1

[Placeholder]

#### Limitation 2

[Placeholder]

### Alternate Approaches


#### Alternate Approach 1

[Placeholder]

#### Alternate Approach 2

[Placeholder]
 
### Potential Future Directions

[Placeholder]

# References
